Fail Faster

### 2025-03-14 paper/paper.tex
more refactoring of rng section

The core of any PBT generator library is a (pseudo) random number generator.
Different PBT libraries use a wide range of different RNGs.
Following the original Haskell QuickCheck, \bq uses the SplitMix algorithm \cn,
implemented (statefully) as an OCaml library called \texttt{Splittable\_random}. Meanwhile,
ScalaCheck uses the JSF algorithm.
% https://www.pcg-random.org/posts/bob-jenkins-small-prng-passes-practrand.html
% https://burtleburtle.net/bob/rand/smallprng.html

@@

Because the RNG sits at the heart of the generation hot path, RNG speed matters enormously.
As we will show in Section~\ref{eval}, speeding up sampling can dramatically speed up generation time and bugfinding speed.

We argue that the RNGs that PBT libraries use are (broadly) a source of inefficiency, and their (lack of) speed is
an impediment towards better bugfinding ability.

Significantly faster RNGs (that still pass pRNG benchmarks like BigCrush) exist like \href{https://github.com/lemire/testingRNG/blob/master/source/lehmer64.h}{this one}, and using them can 
speed up generation tremendously --- in microbenchmarks, a C implementation of the lehmer generator runs 4x faster than a C implementation of splitmix.

Similarly, PBT libraries could even eschew the requirement that a source of entropy pass statistical tests like bigcrush \cn.
This is common practice in testing already: fuzzers often simply use a buffer full of arbitrary bytes as a source of entropy!
Bumping a pointer and reading from a buffer (which can be pipelined trivially) is likely to be significantly faster than any RNG with data-dependent arithmetic instructions.

In the context of this paper, however, demonstrating the benefits of any of the above options is tricky.
Our bug-finding evaluation's power is predicated on the fact that all of the different versions of a generator we test are extensionally identical:
for a given seed and size parameter, they all produce the \emph{same} value.
Using a different random number generator or different source of entropy would break this property.

In Section~\ref{eval}, we demonstrate that faster RNG matters for bugfinding by exploting a ``natural experiment'':
OCaml's \texttt{Splittable\_random} library is slow in a way that can be improved \emph{without} changing its extensional behavior,
namely that it uses \emph{boxed} 64-bit integers. By building a version that uses much faster unoxed 64-bit integer arithmetic, we can
demonstrate just how much bug-finding performance can be improved just by using a more performant RNG.

@@

%  The precise
% details of how SplitMix works are unimportant for the present paper, but the
% main operations are defined in terms of bitwise operations on 64-bit integers.
% However, due to implementation details related to the garbage collector, all
% runtime OCaml values are either (a) pointers to a block of memory, or (b) 63-bit ``immediate'' values.
% As such, the OCaml type of 64-bit ingeters, \texttt{int64}, is \emph{boxed}: it
% is a pointer to a block of memory that contains a 64-bit integer.  This means
% that \emph{all} \texttt{int64} operations (both arithmetic and bitwise) must
% allocate memory cells to contain their output.

>>>>

The core of any PBT generator library is a (pseudo) random number generator.
Different PBT libraries use a wide range of different RNGs.
Following the original Haskell QuickCheck, \bq uses the SplitMix algorithm \cn,
implemented (statefully) as an OCaml library called \texttt{Splittable\_random}. Meanwhile,
ScalaCheck uses the JSF algorithm.

Because the RNG sits at the heart of the generation hot path, RNG speed matters enormously.
We argue that the RNGs that PBT libraries use are (broadly) a source of inefficiency, and their (lack of) speed is
an impediment towards better bugfinding ability \cn \jwc{see comment for cites}.
% https://www.pcg-random.org/posts/bob-jenkins-small-prng-passes-practrand.html
% https://burtleburtle.net/bob/rand/smallprng.html

Significantly faster RNGs like the Lehmer RNG \cn exist like, and using them can
speed up generation tremendously --- in microbenchmarks, a C implementation of
the lehmer generator runs 4x faster than a C implementation of splitmix.
Similarly, PBT libraries could even eschew the requirement that a source of
entropy pass statistical tests\cn.  This is common practice in testing already:
fuzzers often simply use a buffer full of arbitrary bytes as a source of entropy
\cn.  Bumping a pointer and reading from a buffer (which can be pipelined
trivially) will always be significantly faster than any RNG with data-dependent
arithmetic instructions.

In the context of this paper, however, demonstrating the benefits of any of the
above options is tricky.  Our bug-finding evaluation's power is predicated on
the fact that all of the different versions of a generator we test are
extensionally identical: for a given seed and size parameter, they all produce
the \emph{same} value.  Using a different random number generator or different
source of entropy would break this property.  In Section~\ref{eval}, we
demonstrate that faster RNG matters for bugfinding by exploting a ``natural
experiment'': OCaml's \texttt{Splittable\_random} library is slow in a way that
can be improved \emph{without} changing its extensional behavior.
In particular,
due to implementation details related to the OCaml garbage collector,
values of the OCaml type \texttt{int64} are not machine words, but rather
\emph{pointers} to machine words. This means
that \emph{all} \texttt{int64} operations (both arithmetic and bitwise) must
allocate memory cells to contain their output, which has a significant performance benefit.
By building a version that uses much faster ``unboxed'' 64-bit integer arithmetic, we can demonstrate just how much bug-finding
performance can be improved just by using a more performant RNG.

@@

@@

%  The precise
% details of how SplitMix works are unimportant for the present paper, but the
% main operations are defined in terms of bitwise operations on 64-bit integers.
% However,
